---
title: "Ava Hu - Houston Chronicle"
output: html_document
date: "2025-06-03"
---

# How Houston’s Job Market Deviates From the Texas Norm

[avahu\@berkeley.edu](mailto:avahu@berkeley.edu){.email} \| June 2, 2025


### SET UP

```{r}
library(tidyverse)
library(janitor)
library(readxl)
library(rvest)
```

### UPLOAD & CLEAN DATA

#### Houston OES Data

This dataset includes wage and employment information for each occupation in the Houston-The Woodlands-Sugar Land metropolitan area.

```{r}
houston <- read_excel("../data/houston_OES_Report.xlsx", skip = 5) |> 
  clean_names() |> 
  mutate(
    soc_code = str_extract(occupation_soc_code, "\\d{2}-\\d{4}")
  )

```

#### Texas OES Data

This dataset contains overall Texas occupational employment and wage estimates. We'll clean and align it to match the Houston dataset format.#

```{r}
texas <- read_csv("../data/texas_oes.csv") |> 
  clean_names() |> 
  mutate(
    soc_code = str_trim(occ_code)
  )
```

### MERGE & COMPARE
Join the two datasets by SOC code, then calculate wage gaps and prepare for analysis.

```{r}

wage_compare <- inner_join(
  houston, texas,
  by = "soc_code",
  suffix = c("_houston", "_texas")
) |> 
  mutate(
    hourly_mean_houston = as.numeric(hourly_mean_wage),    # from Houston file
    hourly_mean_texas = as.numeric(h_mean),                # from Texas file
    employment_houston = as.numeric(employment_1),
    employment_texas = as.numeric(tot_emp)
  ) |> 
  distinct(soc_code, .keep_all = TRUE)

```

Next, we filtered for occupations where workers in Houston earn less than their counterparts statewide. Then, we calculated the absolute wage gap and sorted the results to find the 10 largest gaps.

```{r}
top10_gap <- wage_compare |> 
  filter(hourly_mean_houston < hourly_mean_texas) |> 
  mutate(
    wage_gap = hourly_mean_texas - hourly_mean_houston
  ) |> 
  arrange(desc(wage_gap)) |> 
  slice(1:10) |> 
  select(
    soc_code,
    occ_title,
    hourly_mean_houston,
    hourly_mean_texas,
    wage_gap,
    employment_houston,
    employment_texas
  )

print(top10_gap)

```

This time, we calculated the reverse: occupations where Houston pays more than the state average. Again, we ranked them by wage difference and kept the top 10.

```{r}
top10_houston_above_texas <- wage_compare |> 
  mutate(wage_gap = hourly_mean_houston - hourly_mean_texas) |> 
  filter(wage_gap > 0) |> 
  arrange(desc(wage_gap)) |> 
  distinct(soc_code, .keep_all = TRUE) |>  
  slice(1:10) |> 
  select(soc_code,
    occ_title,
    hourly_mean_houston,
    hourly_mean_texas,
    wage_gap,
    employment_houston,
    employment_texas)

print(top10_houston_above_texas)

```

### Findings

Several high-skill occupations in Houston show large negative wage gaps compared to the Texas average. Chief executives earn $161.29/hour in Houston, while the statewide average is $349.34, a difference of $188.05. Lawyers make $78.80/hour locally versus $119.67 statewide.

Occupations that earn significantly more in Houston are medical roles. General Internal Medicine Physicians, Psychiatrists, Radiologists, and Pathologists see some of the biggest wage premiums in Houston.Pathologists earn $135.66/hour in Houston, compared to $74.51 statewide. Besides, general and operations managers earn $66.82/hour in Houston versus $37.09 at the state level, with over 100,000 employed locally.

<iframe title="Top 10 Jobs Underpaid in Houston Compared to Texas" aria-label="Range Plot" id="datawrapper-chart-FVp3S" src="https://datawrapper.dwcdn.net/FVp3S/6/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="400" data-external="1"></iframe>

```{=html}
<script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>
```

### Let’s Explore Something Practical: Who’s Doing the Work?

Beyond wage extremes, it's worth asking:
Which low-paid jobs in Houston employ the most people?

I filtered for occupations where Houston wages fall below the statewide average, then ranked by local employment. 

```{r}
top10_low_paid_high_employment <- wage_compare |> 
  filter(hourly_mean_houston < hourly_mean_texas) |> 
  mutate(
    wage_gap = hourly_mean_texas - hourly_mean_houston,
    employment = as.numeric(employment_1)
  ) |> 
  filter(!is.na(employment)) |> 
  arrange(desc(employment)) |> 
  distinct(soc_code, .keep_all = TRUE) |> 
  slice(1:10) |> 
  select(soc_code, occ_title, hourly_mean_houston, hourly_mean_texas, wage_gap, employment)

print(top10_low_paid_high_employment)
```
The result highlights key sectors like education, construction, personal care, and retail—jobs that support the city’s infrastructure and everyday life, yet often offer relatively low pay. Nearly 196,000 people work in education in Houston, earning an average of $31.14 per hour, $15 below the statewide average.

<iframe title="Houston’s Most Common Low-Wage Jobs" aria-label="Table" id="datawrapper-chart-BfgTT" src="https://datawrapper.dwcdn.net/BfgTT/3/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="400" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>

# 5. Houston niche highpay

Top 10 occupations in Houston with higher wages than the Texas average but very low employment

```{r}
top10_high_paid_low_employment <- wage_compare |> 
  filter(hourly_mean_houston > hourly_mean_texas) |> 
  mutate(
    wage_gap = hourly_mean_houston - hourly_mean_texas,
    employment = as.numeric(employment_1)
  ) |> 
  filter(!is.na(employment)) |> 
  arrange(employment) |>  
  distinct(soc_code, .keep_all = TRUE) |> 
  slice(1:10) |> 
  select(soc_code, occ_title, hourly_mean_houston, hourly_mean_texas, wage_gap, employment)

print(top10_high_paid_low_employment)
```

At the same time, many of Houston’s most common jobs — food prep workers, janitors, home health aides — pay below the Texas average. These positions make up a large share of the city’s workforce and are projected to expand in 2025, according to the Greater Houston Partnership’s employment forecast. This suggests a growing reliance on low-wage labor to drive the region’s job growth.

Another focus is the long-term decline of energy jobs in Houston. Oil and gas companies have cut over 60,000 positions in the past decade. Oil remains central to Houston’s economy, but its share of employment is shrinking, replaced by lower-paying sectors like health care, retail, and hospitality. Despite oil production reaching record highs, those gains are no longer tied to job creation.
