---
title: "How Houston’s Job Market Deviates From the Texas Norm"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    self-contained: false
    browser: false
    css: custom.css
date: "2025-06-06"
author: "Ava Hu"
---

Houston is often seen as the oil capital of America, but its job market tells a more complicated story. This analysis shows that while energy jobs are shrinking, lower-paying service sector jobs are growing—and in some cases, dominating the local economy. Many of Houston’s most common occupations, like fast food workers and cashiers, earn less than the statewide average, even as they support the city’s daily life. Meanwhile, some of the highest-paid roles—like podiatrists and internal medicine physicians—are vanishingly rare.

This imbalance raises a key question: Is Houston becoming a city of low-wage, high-volume work?

My analysis focuses on:

-   **Which sectors are growing or shrinking in Houston?**\
    Using monthly data from 2015 to 2024, I calculate annual average employment and show which sectors are driving growth.

-   **Is Houston moving away from oil?**\
    By comparing trends in oil-linked sectors with service-oriented industries, I assess whether the city is undergoing a structural shift.

-   **Which occupations in Houston are underpaid compared to the Texas average?**\
    I identify jobs where workers in Houston earn less than their statewide counterparts—and how widespread those roles are.

-   **Which low-wage jobs employ the most people in Houston?**\
    Some of the city’s most common jobs, such as food service or customer support, also offer wages below the state average.

-   **Are there high-paying jobs that are rare in Houston?**\
    I rank the most well-compensated roles that have surprisingly low local employment.

## Which Sectors Are Growing in Houston?


<iframe title="Percentage Change in Employment by Sector " aria-label="Interactive line chart" id="datawrapper-chart-RNfkh" src="https://datawrapper.dwcdn.net/RNfkh/1/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="471" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>


<iframe title="Energy Sector Employment in Houston" aria-label="Interactive line chart" id="datawrapper-chart-zqLUq" src="https://datawrapper.dwcdn.net/zqLUq/1/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="466" data-external="1">

</iframe>

```{r, results='asis', echo=FALSE}
cat('
<script type="text/javascript">
!function(){"use strict";window.addEventListener("message",(function(a){
  if(void 0!==a.data["datawrapper-height"]){
    var e=document.querySelectorAll("iframe");
    for(var t in a.data["datawrapper-height"])
      for(var r,i=0;r=e[i];i++)
        if(r.contentWindow===a.source){
          var d=a.data["datawrapper-height"][t]+"px";
          r.style.height=d
        }
  }
}))}();
</script>
')
```

```{r, results='asis', echo=FALSE}
cat('
<iframe src="https://datawrapper.dwcdn.net/FVp3S/6/" style="width:100%; height:400px; border:none;" frameborder="0" scrolling="no"></iframe>
')
```

```{r, results='asis', echo=FALSE}
cat('
<iframe src="https://datawrapper.dwcdn.net/BfgTT/3/" style="width:100%; height:400px; border:none;" frameborder="0" scrolling="no"></iframe>
')

```

```{r, results='asis', echo=FALSE}
cat('
<script type="text/javascript">
!function(){"use strict";window.addEventListener("message",(function(a){
  if(void 0!==a.data["datawrapper-height"]){
    var e=document.querySelectorAll("iframe");
    for(var t in a.data["datawrapper-height"])
      for(var r,i=0;r=e[i];i++)
        if(r.contentWindow===a.source){
          var d=a.data["datawrapper-height"][t]+"px";
          r.style.height=d
        }
  }
}))}();
</script>
')
```

## R Markdown

#### The Data

**1-Texas_OES_Report**: Based on the U.S. Bureau of Labor Statistics’ May 2024 Occupational Employment and Wage Statistics. I used statewide data across all sectors. Source: [BLS OES Research Estimates](https://www.bls.gov/oes/2024/may/oes_research_estimates.htm).

**2-Houston_OES_Report**: Derived from the May 2024 OES data published by the U.S. Bureau of Labor Statistics. I downloaded the data for the Houston-Pasadena-The Woodlands, TX metro area from the [Metropolitan and Nonmetropolitan Area Table](https://www.bls.gov/oes/2024/may/oessrcma.htm#T).

**3-SeriesReport**: Based on the U.S. Bureau of Labor Statistics’ [State and Area Employment, Hours, and Earnings](https://www.bls.gov/data/#employment) (SAE) dataset. This time series data provides monthly employment figures by industry sector for specific metropolitan areas. I downloaded separate reports for five sectors within the Houston-The Woodlands-Sugar Land, TX MSA. For each sector, I used non-seasonally adjusted monthly data from 2015 to 2024.

### Set Up

```{r}
library(tidyverse)
library(janitor)
library(readxl)
library(rvest)
```

## Which Sectors Are Growing in Houston? (2015–2024)

To understand long-term employment trends, I wanted a clearer picture than a single month could provide. So I calculated the annual average employment for each sector, using monthly data from January through December.

I pulled in five key sectors that identified above: Mining, Logging, and Construction; Service-Providing; Leisure and Hospitality; Retail Trade and Education and Health Services from the Bureau of Labor Statistics, cleaned the data, and averaged the twelve monthly values for each year from 2015 to 2024.

The result is a table that shows, year by year, how many people were employed in each sector on average—giving a more stable view of how Houston’s job market is shifting across industries.

```{r}
# Set your data folder path
data_path <- "~/Documents/GitHub/ava-hu/data"

# Only load the 5 sector files
file_paths <- list.files(
  path = data_path,
  pattern = "^SeriesReport.*\\.xlsx$|^HoustonMining.*\\.xlsx$",
  full.names = TRUE
)

# Get sector name from metadata
get_sector_name <- function(file) {
  meta <- read_excel(file, range = "A8:B8", col_names = FALSE)
  sector <- meta[[2]]
  if (is.na(sector)) sector <- str_remove(basename(file), "\\.xlsx")
  return(sector)
}

# Extract and compute Jan–Dec average for each year
extract_annual_avg <- function(file) {
  df <- read_excel(file, skip = 12) |> 
    clean_names() |> 
    select(year, jan:dec) |> 
    filter(!is.na(year) & str_detect(as.character(year), "^\\d{4}")) |> 
    mutate(across(jan:dec, as.numeric)) |> 
    mutate(year = as.integer(year))

  df$annual_avg <- rowMeans(df[, c("jan", "feb", "mar", "apr", "may", "jun",
                                   "jul", "aug", "sep", "oct", "nov", "dec")],
                            na.rm = TRUE)

  df <- df |> 
    filter(year >= 2015, year <= 2024)

  sector <- get_sector_name(file)

  df |> 
    select(year, annual_avg) |> 
    pivot_wider(names_from = year, values_from = annual_avg) |> 
    mutate(Sector = sector) |> 
    select(Sector, everything())
}

# Run across all files
sector_avg_table <- bind_rows(lapply(file_paths, extract_annual_avg))

# View result
print(sector_avg_table)

```

```{r}
trend_summary <- sector_avg_table |> 
  pivot_longer(-Sector, names_to = "year", values_to = "avg_employment") |> 
  mutate(
    year = as.integer(year),
    industry = Sector
  ) |> 
  select(industry, year, avg_employment)

trend_pct <- trend_summary |> 
  group_by(industry) |> 
  mutate(
    base_2015 = avg_employment[year == 2015],
    pct_change = (avg_employment - base_2015) / base_2015 * 100
  )

# Label for 2024
label_2024 <- trend_pct |> 
  filter(year == 2024) |> 
  mutate(
    label = paste0(
      ifelse(pct_change >= 0, "+", ""),
      round(pct_change, 1), "%"
    )
  )

# Plot with Y axis in %
ggplot(trend_pct, aes(x = year, y = pct_change, color = industry)) +
  geom_line(size = 1.1) +
  geom_text(data = label_2024, aes(label = label), 
            hjust = -0.1, vjust = 0.5, show.legend = FALSE, size = 3.5) +
  xlim(2015, 2025.5) +
  labs(
    title = "Percent Change in Employment by Sector (Relative to 2015)",
    subtitle = "Annual average employment, 2015–2024",
    x = "Year", y = "Change Since 2015 (%)",
    color = "Sector"
  ) +
  theme_minimal()

```

## Is Houston moving away from Oil?

Although Houston has long been known as an oil capital, the city’s job market is shifting away from its energy roots. Between 2015 and 2024, employment in Mining, Logging, and Construction fell by 1.9%, making it the only major sector to experience a net decline over the decade. In contrast, service-oriented industries have seen substantial growth.

```         
All Employees, In Thousands
```

```{r}
data_path <- "~/Documents/GitHub/ava-hu/data/oil"
files <- list.files(data_path, pattern = "\\.xlsx$", full.names = TRUE)

read_bls_file_with_label <- function(file) {
  
  industry_name <- read_excel(file, sheet = "BLS Data Series", range = "B9", col_names = FALSE) |> pull(1)

  read_excel(file, sheet = "BLS Data Series", skip = 11) |>
    rename(Year = 1) |>
    pivot_longer(-Year, names_to = "Month", values_to = "Employment") |>
    filter(!is.na(Employment)) |>
    mutate(
      Industry = industry_name,
      date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%b-%d")
    )
}

oil_data <- map_dfr(files, read_bls_file_with_label)

```

```{r}
oil_data |>
  mutate(Year = lubridate::year(date)) |>
  group_by(Industry, Year) |>
  summarise(annual_avg = mean(Employment, na.rm = TRUE), .groups = "drop") |>
  arrange(Industry, Year) |>
  group_by(Industry) |>
  mutate(change = annual_avg - lag(annual_avg)) |>
  filter(!is.na(change)) |>
  top_n(-1, change)  


```

### UPLOAD & CLEAN DATA

#### Houston OES Data

This dataset includes wage and employment information for each occupation in the Houston-The Woodlands-Sugar Land metropolitan area.

```{r}
houston_clean <- read_excel("../data/houston_OES_Report.xlsx", skip = 5) |> 
  clean_names() |> 
  mutate(
    soc_code = str_extract(occupation_soc_code, "\\d{2}-\\d{4}"),
    hourly_mean_houston = as.numeric(hourly_mean_wage),
    employment_houston = as.numeric(employment_1)
  ) |> 
  filter(!is.na(soc_code), !is.na(hourly_mean_houston), !is.na(employment_houston))

```

#### Texas OES Data

This dataset contains overall Texas occupational employment and wage estimates. We'll clean and align it to match the Houston dataset format.

To account for the fact that Texas OES data includes multiple entries per occupation across different industries (NAICS classifications), I grouped the dataset by SOC code. For each occupation, I calculated the average hourly wage across all industries and summed up the total employment. This aggregation provides a single representative wage and employment figure per occupation, allowing a clean, one-to-one comparison with the Houston dataset.

```{r}
texas <- read_csv("../data/texas_oes.csv") |> 
  clean_names() |> 
  mutate(
    soc_code = str_trim(occ_code)
  )

texas_clean <- texas |> 
  clean_names() |> 
  mutate(soc_code = str_trim(occ_code)) |> 
  group_by(soc_code) |> 
  summarise(
    hourly_mean_texas = mean(as.numeric(h_mean), na.rm = TRUE),
    employment_texas = sum(as.numeric(tot_emp), na.rm = TRUE),
    .groups = "drop"
  )
```

### MERGE & COMPARE

Join the two datasets by SOC code, then calculate wage gaps and prepare for analysis.

```{r}

wage_compare <- inner_join(
  houston_clean, texas_clean,
  by = "soc_code"
) |> 
  distinct(soc_code, .keep_all = TRUE)

```

Next, we filtered for occupations where workers in Houston earn less than their counterparts statewide. Then, we calculated the absolute wage gap and sorted the results to find the 10 largest gaps.

```{r}

top10_low_paid_high_employment <- wage_compare |> 
  filter(hourly_mean_houston < hourly_mean_texas) |> 
  mutate(
    wage_gap = hourly_mean_texas - hourly_mean_houston
  ) |> 
  arrange(desc(employment_houston)) |> 
  slice(1:10) |> 
  select(
    soc_code,
    occ_title = occupation_soc_code,
    hourly_mean_houston,
    hourly_mean_texas,
    wage_gap,
    employment_houston
  )

print(top10_low_paid_high_employment)


```

This time, we calculated the reverse: occupations where Houston pays more than the state average. Again, we ranked them by wage difference and kept the top 10.

```{r}
top10_houston_above_texas <- wage_compare |> 
  filter(hourly_mean_houston > hourly_mean_texas) |> 
  mutate(wage_gap = hourly_mean_houston - hourly_mean_texas) |> 
  arrange(desc(wage_gap)) |> 
  distinct(soc_code, .keep_all = TRUE) |>  
  slice(1:10) |> 
  select(
    soc_code,
    occ_title = occupation_soc_code,
    hourly_mean_houston,
    hourly_mean_texas,
    wage_gap,
    employment_houston,
    employment_texas
  )

print(top10_houston_above_texas)


```

### Findings

The top 10 underpaid jobs in Houston show relatively small wage differences, typically less than \$3/hour, with the largest gap seen in sales and related occupations, where Houston workers earn about \$7.36/hour less than their statewide counterparts. The rest of the list includes fast food workers, construction laborers, cashiers, stockers, and customer service representatives, all of whom experience minor pay disadvantages compared to the Texas average.

Occupations that earn significantly more in Houston are medical roles.At the top of the list are pathologists, who earn \$135.66/hour in Houston—nearly \$47/hour more than the statewide average. Other medical professions such as ophthalmologists, internal medicine physicians, dentists, and podiatrists also earn over \$20/hour more in Houston than in Texas overall.

### Which Low-Wage Jobs Employ the Most People in Houston?

Beyond wage extremes, it's worth asking: Which low-paid jobs in Houston employ the most people?

I filtered for occupations where Houston wages fall below the statewide average, then ranked by local employment.

```{r}
top10_low_paid_high_employment <- wage_compare |> 
  filter(hourly_mean_houston < hourly_mean_texas) |> 
  mutate(
    wage_gap = hourly_mean_texas - hourly_mean_houston
  ) |> 
  arrange(desc(employment_houston)) |> 
  distinct(soc_code, .keep_all = TRUE) |> 
  slice(1:10) |> 
  select(
    soc_code,
    occ_title = occupation_soc_code,
    hourly_mean_houston,
    hourly_mean_texas,
    wage_gap,
    employment_houston
  )

print(top10_low_paid_high_employment)

```

The result highlights key sectors like education, construction, personal care, and retail—jobs that support the city’s infrastructure and everyday life, yet often offer relatively low pay. Nearly 196,000 people work in education in Houston, earning an average of \$31.14 per hour, \$15 below the statewide average.

### Top 10 occupations in Houston with higher wages than the Texas average but very low employment

I also looked for occupations where the average hourly wage in Houston is higher than the statewide average, then ranked them by local employment. From there, I selected the 10 least common roles.

```{r}

top10_high_paid_low_employment <- wage_compare |> 
  filter(hourly_mean_houston > hourly_mean_texas) |> 
  mutate(
    wage_gap = hourly_mean_houston - hourly_mean_texas
  ) |> 
  arrange(employment_houston) |>  
  distinct(soc_code, .keep_all = TRUE) |> 
  slice(1:10) |> 
  select(
    soc_code,
    occ_title = occupation_soc_code,
    hourly_mean_houston,
    hourly_mean_texas,
    wage_gap,
    employment_houston
  )

print(top10_high_paid_low_employment)

```

Some of the best-paying opportunities in Houston are highly specialized and relatively rare. For instance, podiatrists in Houston earn nearly \$111/hour, about \$22 more than the statewide average, but only 70 people work in that area.
